Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2008-06-14 SCP2008/paper/metamonads.tex
various small changes throughout the text

let ge = fun a_orig ->
    let r = ref 0 in                    (* current row index, 0-based *)
    let c = ref 0 in                    (* current column index, 0-based *)
    let a = Array.copy (a_orig.arr) in  (* to avoid clobbering A, save it *)
    let m = a_orig.m in                 (* the number of columns *)
    let n = a_orig.n in                 (* the number of rows *)
    let det_sign = ref 1 in             (* Accumulate sign and magnitude *)
    let det_magn = ref 1 in             (*   of the determinant *)
    while !c < m && !r < n do
        (* Look for a pivot *)
        let pivot = find_pivot a (n,m) !r !c in
        let piv_val = (match  pivot with
        | Some ((piv_r, piv_c),piv_val) ->
            if piv_c <> !c then
               begin
                 swap_cols a (n,m) !c piv_c;
                 det_sign := - !det_sign (* flip the sign of the det *)
               end;
            if piv_r <> !r then
               begin
                 swap_rows a (n,m) (!r,!c) piv_r;
                 det_sign := - !det_sign (* flip the sign of the det *)
               end;
            Some piv_val
        | None -> None) in
        (* now do the row-reduction over the (r,c)-(n,m) block *)
        (match piv_val with
        | Some a_rc -> begin
            for ii = !r+1 to n-1 do
                let cur = a.(ii*m + !c) in
                if not (cur == 0)  then
                  begin
                    for j = !c+1 to m-1 do
                      (* fraction-free elimination *)
                      a.(ii*m+j) <- (a.(ii*m+j) * a_rc - a.(!r*m+j) * cur) 
                                    / ! det_magn
                    done;
                    a.(ii*m+ !c) <- 0
                  end;
            done;
            det_magn := a_rc;
            r := !r + 1                 (* advance the rank only if pivot > 0*)
          end
        | None -> det_sign := 0);
        c := !c + 1
    done;
    (* Final result *)
    ({arr=a; n=n; m=m},                 (* The matrix now has the U factor*)
    (if      !det_sign = 0 then  0      (* Compute the signed det *)
     else if !det_sign = 1 then  !det_magn 
     else                        (- !det_magn)),
    !r)                                 (* Rank *)
\end{code}
\caption{Fraction-free in-place Gaussian-elimination over an integer
  matrix}\label{fig:GE-algorithm}
\end{figure}
As typically presented in the textbooks, the main algorithm depends on
the separately-defined function |pivot| to find a pivot. We use
full-pivoting, i.e., searching the non-yet examined portion of the matrix, 
the rectangular block |(r,c)-(n,m)|, for the element with the 
maximum absolute value. The function returns the value of the
found pivot and its location; the return value is an option type with
|None| indicating that all examined elements are zeros. In the main
algorithm, after we found the pivot, we swap the current column 
with the pivot column, and swap the current row with the pivot row. 
After the swap, the $a_{rc}$ element of the matrix is the pivot.
Swapping two rows or two columns changes the sign of the determinant.
After the swapping, the algorithm does so-called row-reduction
over the |(r,c)-(n,m)| block of the matrix containing not-yet
processed elements. We implement the \emph{fraction-free}
algorithm \oleg{Citation needed!}:
the division operation in line XXX assuredly divides two integers
with no remainder, which requires the accumulation
of the determinant .

@@

In this section, we use the the sample code in
Fig.~\ref{fig:GE-algorithm} to identify (some of the) aspects
discussed in \S\ref{sec:aspects}. We abstract these aspects and
describe their implementation as modules of the code generator.  We
then present the generic Gaussian Elimination algorithm -- what is
left after all the aspects are abstracted away. Finally, we show an
instantiation of the generic generator, whose execution yields code
for the algorithm of Fig.~\ref{fig:GE-algorithm}, this time
automatically generated rather than manually written.

@@

\subsection{Abstract code and lifted domains}\label{sec:lifteddomains}

@@

The interpretation aspect is too implemented as an OCaml module.  The
aspect is quite large -- there are many basic operations to abstract
over -- and so the module is structured into several
sub-modules. First we introduce an abstract type |('a,'b) rep| which
describes what sort of objects the interpretation may produce (e.g.,
strings with C code, AST, etc). The type has two parameters: the
second specifies the type of the object, and the first is the
`placeholder' for all other information that may need to be tracked
about the object in a particular interpretation. The interpretation as
MetaOCaml code values uses the first parameter of |rep| to track the
environment classifier.
\oleg{I'm describing in this section coderep.mli and one of its
  instances, code.ml. The only change is that I use ('a,'b) rep rather
  than ('a,'b) abstract, for uniformity and clarity. Why did we even
  need to introduce this additional distinction, `rep' vs
  `abstract'. Also, the name `abstract' is a bit too
  general. Anyway, the explanation does not have to follow the code to
  the letter, as Reviewer 2 has suggested.}

@@

As we argued in \S\ref{CPS}, we will be writing our generic GE
algorithm in the monadic style. For convenience we define the
following type synonyms for our |('p,'v) monad| of Fig.\ref{ourmonad}:
The first is used when we are always generating (abstract) code, while
the second helps when we might generate (abstract) code. The helper
type synonym |('pc,'p,'a) cmonad_constraint| is effectively the
abbreviation for a set of constraints imposed on its arguments.
\begin{code}
type ('pc,'p,'a) cmonad_constraint = unit
      constraint
	  'p = <state : 's list; answer : ('a,'w) abstract>
      constraint
          'pc = <classif : 'a; answer : 'w; state : 's; ..>

type ('pc,'v) cmonad = ('p,('a,'v) abstract) monad
      constraint _ = ('pc,'p,'a) cmonad_constraint

type ('pc,'v) omonad = ('p,('a,'v) abstract option) monad
      constraint _ = ('pc,'p,'a) cmonad_constraint 
\end{code}

@@

We will consider the interpretation aspect ``open'' and so we can
refer to all operations such as |apply|, |seqM|, etc, without further
qualification.

@@

The signature |CONTAINER2D| below specifies that a container must
provide functions |dim1| and |dim2| to extract the dimensions,
functions |getL| to generate container getters, the cloning generator
|copy| and functions that generate code for row and column
swapping. The inclusion of these functions in the signature of all
containers makes it simpler to optimize the relevant functions
depending on the actual representation of the container while not
burdening the users of containers with efficiency details (see
\S\ref{GE-algorithm} for an example of such an optimization).  We
could have made the type |CONTAINER2D| a functor, parameterized over
|DOMAINL|. The current design simplifies the signature of the generic
GE algorithm at a cost of an extra instantiation step (The
instantiation of the generic algorithm is described in
\S\ref{sec:main-generation}. The expense is worthwhile: currently
OCaml is quite slow when processing functors with complex signatures).
\begin{code}
module type CONTAINER2D = sig
  module Dom:DOMAINL
  type contr
  type 'a vc = ('a,contr) rep
  type 'a vo = ('a,Dom.v) rep
  val getL : 'a vc -> ('a,int) rep -> ('a,int) rep -> 'a vo
  val dim1 : 'a vc -> ('a,int) rep
  val dim2 : 'a vc -> ('a,int) rep
  val mapper : ('a vo -> 'a vo) option -> 'a vc -> 'a vc
  val copy : 'a vc -> 'a vc
  val init : ('a,int) rep -> ('a, int) rep -> 'a vc
  val augment : 'a vc -> ('a,int) rep -> ('a, int) rep -> 'a vc ->
                ('a, int) rep -> 'a vc
  val identity : ('a,int) rep -> ('a, int) rep -> 'a vc
  val swap_rows_stmt : 'a vc -> ('a, int) rep -> ('a, int) rep -> 
                       ('a,unit) rep
  val swap_cols_stmt : 'a vc -> ('a, int) rep -> ('a, int) rep -> 
                       ('a,unit) rep
  val row_head : 'a vc -> ('a, int) rep -> ('a, int) rep -> 'a vo
  val col_head_set : 'a vc -> ('a,int) rep -> ('a,int) rep -> 'a vo -> 
            ('a,unit) rep
end
\end{code}

The type of our containers include the lifted domain |Dom| as one of
the components. This is quite convenient since operations on containers
are usually accompanied by the operations on the retrieved and stored
values. Our containers are parametric over a |DOMAINL|, i.e., functors
from a |DOMAINL| module to the actual implementation of a
container. For example, the following functor defines a matrix
container as a single array, with elements stored in the
row-major order -- the container used in Fig.\ref{fig:GE-algorithm}.
\begin{code}[commandchars=\\\{\}]
module GenericVectorContainer(Dom:DOMAINL) =
  struct
  module Dom = Dom
  type contr = Dom.v container2dfromvector
  type 'a vc = ('a,contr) code
  type 'a vo = ('a,Dom.v) code
  let getL x i j = .< ((.~x).arr).(.~i* (.~x).m + .~j) >.
  let dim2 x = .< (.~x).n >.      (* number of rows *)
  let dim1 x = .< (.~x).m >.      (* number of cols *)
  let mapper g a = match g with
  \dots
\end{code}
%
The accompanying code \cite{metamonadsURL} includes containers whose
elements are stored in a 1D array in a column-wise (Fortran-like)
modes, and container for a matrix represented as arrays of rows.

The generic GE algorithm described later is a functor over the
container.  Given a particular container aspect, the functor yields a
module containing various algorithmic aspects for the user to choose,
as well as the main GE driver. The given container instance is
available to all these aspects under the name of |C| (so that the type
of the container can be referred to as |C.contr| and the dimensions
can be obtained using |C.dim1| and |C.dim2|). The numeric domain, which
is part of the container, can be referred to as |D.Dom| and the type
of the domain elements is |C.Dom.v|. We shall see many such references
as we describe particular algorithmic aspects below.

@@

The determinant aspect is one of several tracking and strategic
aspects. The main function of the algorithm invokes the functions of
these aspects at some interesting points during the computation (e.g.,
pivot is required, two rows have to be permuted, the final answer to the
user has to be built). The instance of the aspect may do something (or
nothing) at each point, for example, find pivot according to the
pivoting strategy (such as row pivoting, any non-zero element), update
the current value of the determinant, etc.

Our sample Gaussian Elimination algorithm, Fig.\ref{fig:GE-algorithm},
demonstrates that tracking determinant is complex: first we define the
variables |det_sign| and |det_magn| used for tracking (lines 22 and
23), then we have to flip the sign when swapping two rows or two
columns or when the matrix found to be singular (lines 32, 37, 59).
The value of the determinant should be updated per each pivoting (line
56). Finally, we convert the tracking state to the resulting
determinant value (lines 64-66). We observe these lines are not
contiguous: the determinant aspect is consulted at several separate
places in the GE algorithm, and the aspect is supposed to keep
state. As other aspects, the determinant aspect is a module of the
following signature:
\begin{code}
module type DETERMINANT = sig
  type tdet = C.Dom.v ref
  type 'a lstate
  type 'pc pc_constraint = unit
    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
  type ('pc,'v) lm = ('pc,'v) cmonad
    constraint _  = 'pc pc_constraint
  type ('pc,'v) om = ('pc,'v) omonad
    constraint _  = 'pc pc_constraint
  type 'pc nm = ('p,unit) monad
    constraint _ = ('pc,'p,_) cmonad_constraint
    constraint _ = 'pc pc_constraint
  val decl : unit -> 'b nm (* no code is generated *)
  val upd_sign  : unit -> ('b,unit) om
  val zero_sign : unit -> ('b,unit) lm
  val acc_magn  : ('a,C.Dom.v) abstract -> (<classif : 'a; ..>,unit) lm
  val get_magn  : unit -> ('b,tdet) lm
  val set_magn  : ('a,C.Dom.v) abstract -> (<classif : 'a; ..>,unit) lm
  val fin       : unit -> ('b,C.Dom.v) lm
end
\end{code}
The type |'a lstate| denotes the state being kept as part of the
overall monadic state (tagged as |`TDet|). The signature specifies the
operations of the aspect, such as |decl| for initializing the tracking
state, |zero_sign| to note matrix singularity, |set_magn| to set the
magnitude and |fin| to convert the tracking state to the resulting
value, of  the type |C.Dom.v|, the type of the container
elements. Other aspects follow a similar outline.
\oleg{The following is perhaps too particular to mention in the
  paper. AT least we can keep it in comments:
  We stress the difference in return types of \texttt{decl},
  \texttt{upd\_sign} and \texttt{zero\_sign} functions. The former says
  that \texttt{decl} is an action which is executed only for its
  side-effect. It yields no interpretation object (e.g., produces no
  code value). The function \texttt{zero\_sign} always produces a code
  value -- an expression such as assignment that has the type unit. 
  The function
  \texttt{upd\_sign} is the generator that may produce code, or may
  not. The option type lets us avoid generating code such as 
  \texttt{a := b; ()} with useless dummies.}

@@

For completeness, we briefly describe and show the signatures of the
other aspects. The |RANK| aspect tracks rank; the aspect is a simpler
version of the determinant tracking aspect.  |PIVOTKIND| describes the
representation of the permutation matrix to track permutations
executed in pivoting; |TRACKPIVOT| specifies if these permutations are
tracked at all. |LOWER| tracks the |L| factor when GE algorithm is
used for the in-place LU-decomposition. |PIVOT| is a higher-order
functor abstracting over the pivoting algorithm (full pivoting, row
pivoting, no pivoting, etc). If pivot is found, the aspect should swap
rows and columns appropriately. The swapping may need to be tracked if
we wish to return the permutation matrix. If the determinant is
tracked, the swapping must update the sign of the determinant. The
wapiing also affects the L factor being accumulated in-place. The
|UPDATE| aspect abstracts over the algorithm for updating matrix
elements during row-reductions: full division update or fraction-free
update. Finally, the |INPUT| aspect is an interface aspect abstracting
over the choice of augmented matrix.

\begin{code2}
module type RANK = sig
  type 'a tag_lstate = 'a tag_lstate_
  val decl   : unit -> ('b, int ref) lm
  val succ   : unit -> ('b, unit) lm
  val fin    : unit -> ('b, int) lm
end

@@

The type |wmatrix| is for convenience, a bundling of several pieces
of data which are needed by both the |OUTPUT| aspect and the |PIVOT| 
aspect.  It is defined as
\begin{code}
type 'a wmatrix = {matrix: 'a C.vc; numrow: ('a,int) rep; 
                   numcol: ('a,int) rep}
\end{code}

The module of signature |INTERNAL_FEATURES| bundles information
tracking aspects. The latter are not directly selectable by the
user. Rather, they are \emph{functions} of other user choices. The
implementations of the tracking aspects such as |Rank|, |NoRank|,
|PackedLower| are quite similar (and simpler) than the implementation
of the |AbstractDet| and |NoDet| aspects in \S\ref{sec:det}. The
choice of a particular tracking aspect may depend on all other choices
(e.g., it is not possible to extract the |L| factor if the domain is
not a field). Currently we use preflight checks to ensure consistency
of tracking aspects. Previously~\cite{CaretteKiselyov05} we tried to
implement all our preflight tests at the (module) type level, using
sharing constraints and module computations. That lead to obscure
code, long impenetrable type error messages and made compilation very
slow.  Furthermore, at least in OCaml, type-level computations are not
powerful enough to help us with tasks such as verifying that an integer
is prime, as required in \S\ref{sec:domains}.

>>>>

let ge = fun a_orig ->
    let r = ref 0 in                    (* current row index, 0-based *)
    let c = ref 0 in                    (* current column index, 0-based *)
    let a = Array.copy (a_orig.arr) in  (* to avoid clobbering A, save it *)
    let m = a_orig.m in                 (* the number of columns *)
    let n = a_orig.n in                 (* the number of rows *)
    let det_sign = ref 1 in             (* Accumulate sign and magnitude *)
    let det_magn = ref 1 in             (*   of the determinant *)
    while !c < m && !r < n do
        (* Look for a pivot *)
        let pivot = find_pivot a (n,m) !r !c in
        let piv_val = (match  pivot with
        | Some ((piv_r, piv_c),piv_val) ->
            if piv_c <> !c then
               begin
                 swap_cols a (n,m) !c piv_c;
                 det_sign := - !det_sign (* flip the sign of the det *)
               end;
            if piv_r <> !r then
               begin
                 swap_rows a (n,m) (!r,!c) piv_r;
                 det_sign := - !det_sign (* flip the sign of the det *)
               end;
            Some piv_val
        | None -> None) in
        (* now do the row-reduction over the (r,c)-(n,m) block *)
        (match piv_val with
        | Some a_rc -> begin
            for ii = !r+1 to n-1 do
                let cur = a.(ii*m + !c) in
                if not (cur == 0)  then
                  begin
                    for j = !c+1 to m-1 do
                      (* fraction-free elimination *)
                      a.(ii*m+j) <- (a.(ii*m+j) * a_rc - a.(!r*m+j) * cur) 
                                    / ! det_magn
                    done;
                    a.(ii*m+ !c) <- 0
                  end;
            done;
            det_magn := a_rc;
            r := !r + 1                 (* advance the rank only if pivot > 0*)
          end
        | None -> det_sign := 0);
        c := !c + 1
    done;
    (* Final result *)
    ({arr=a; n=n; m=m},                 (* The matrix now has the U factor*)
    (if      !det_sign = 0 then  0      (* Compute the signed det *)
     else if !det_sign = 1 then  !det_magn 
     else                        (- !det_magn)),
    !r)                                 (* Rank *)
\end{code}
\caption{Fraction-free in-place Gaussian-elimination over an integer
  matrix}\label{fig:GE-algorithm}
\end{figure}
As typically presented in the textbooks, the main algorithm depends on
the separately-defined function |pivot| to find a pivot. We use
full-pivoting, i.e., searching the non-yet examined portion of the matrix, 
the rectangular block |(r,c)-(n,m)|, for the element with the 
maximum absolute value. The function returns the value of the
found pivot and its location; the return value is an option type with
|None| indicating that all examined elements are zeros. In the main
algorithm, after we found the pivot, we swap the current column 
with the pivot column, and swap the current row with the pivot row. 
After the swap, the $a_{rc}$ element of the matrix is the pivot.
Swapping two rows or two columns changes the sign of the determinant.
After the swapping, the algorithm does so-called row-reduction
over the |(r,c)-(n,m)| block of the matrix containing not-yet
processed elements. We implement the \emph{fraction-free}
algorithm \oleg{Citation needed!}:
the division operation in line 51 assuredly divides two integers
with no remainder, which requires the accumulation
of the determinant.

@@

In this section, we use the the sample code in
Fig.~\ref{fig:GE-algorithm} to identify (some of the) aspects
discussed in \S\ref{sec:aspects}. We abstract these aspects and
describe their implementation as modules of the code generator.  We
then present in \S\ref{sec:main-generation} the generic Gaussian
Elimination algorithm -- what is left after all the aspects are
abstracted away. Finally, we show an instantiation of the generic
generator, whose execution yields code for the algorithm of
Fig.~\ref{fig:GE-algorithm}, this time automatically generated rather
than manually written.

@@

\subsection{Abstracting interpretations}\label{sec:lifteddomains}

@@

The interpretation aspect is too implemented as an OCaml module.  The
aspect is quite large -- there are many basic operations to abstract
over -- and so the module is structured into several
sub-modules. First we introduce an abstract type |('a,'b) rep| which
describes what sort of objects the interpretation may produce (e.g.,
AST, strings with C code, etc). The type has two parameters: the
second specifies the type of the object, and the first is the
`placeholder' for all other information that may need to be tracked
about the object in a particular interpretation. The interpretation as
MetaOCaml code values uses the first parameter of |rep| to track the
environment classifier.
\oleg{I'm describing in this section coderep.mli and one of its
  instances, code.ml. The only change is that I use ('a,'b) rep rather
  than ('a,'b) abstract, for uniformity and clarity. Why did we even
  need to introduce this additional distinction, `rep' vs
  `abstract'. Also, the name `abstract' is a bit too
  general. Anyway, the explanation does not have to follow the code to
  the letter, as Reviewer 2 has suggested.}

@@

As we argued in \S\ref{CPS}, we will be writing our generic GE
algorithm in the monadic style. For convenience we define the
following type synonyms for our |('p,'v) monad| of Fig.~\ref{ourmonad}:
The first is used for a monadic action that always produces a useful
value; the |omonad| synonym describes a monadic action that may
produce an interpretation object. The helper
type synonym |('pc,'p,'a) cmonad_constraint| is effectively the
abbreviation for a set of constraints imposed on its arguments.
\begin{code}
type ('pc,'p,'a) cmonad_constraint = unit
  constraint 'p = <state : 's list; answer : ('a,'w) rep>
  constraint 'pc = <classif : 'a; answer : 'w; state : 's; ..>

type ('pc,'v) cmonad = ('p,('a,'v) rep) monad
  constraint _ = ('pc,'p,'a) cmonad_constraint

type ('pc,'v) omonad = ('p,('a,'v) rep option) monad
  constraint _ = ('pc,'p,'a) cmonad_constraint 
\end{code}

@@

We consider the interpretation aspect ``open'' and so we can
refer to all operations such as |apply|, |seqM|, etc, without further
qualification.

@@

The signature |CONTAINER2D| below specifies that a container must
provide functions |dim1| and |dim2| to extract the dimensions,
functions |getL| to generate container getters, the cloning generator
|copy| and functions that generate code for row and column
swapping. The inclusion of these functions in the signature of all
containers makes it simpler to optimize the relevant functions
depending on the actual representation of the container while not
burdening the users of containers with efficiency details (see
\S\ref{GE-algorithm} for an example of such an optimization).
\begin{code}
module type CONTAINER2D = sig
  module Dom:DOMAINL
  type contr
  type 'a vc = ('a,contr) rep
  type 'a vo = ('a,Dom.v) rep
  val getL : 'a vc -> ('a,int) rep -> ('a,int) rep -> 'a vo
  val dim1 : 'a vc -> ('a,int) rep
  val dim2 : 'a vc -> ('a,int) rep
  val mapper : ('a vo -> 'a vo) option -> 'a vc -> 'a vc
  val copy : 'a vc -> 'a vc
  val init : ('a,int) rep -> ('a, int) rep -> 'a vc
  val augment : 'a vc -> ('a,int) rep -> ('a, int) rep -> 'a vc ->
                ('a, int) rep -> 'a vc
  val identity : ('a,int) rep -> ('a, int) rep -> 'a vc
  val swap_rows_stmt : 'a vc -> ('a, int) rep -> ('a, int) rep -> 
                       ('a,unit) rep
  val swap_cols_stmt : 'a vc -> ('a, int) rep -> ('a, int) rep -> 
                       ('a,unit) rep
  val row_head : 'a vc -> ('a, int) rep -> ('a, int) rep -> 'a vo
  val col_head_set : 'a vc -> ('a,int) rep -> ('a,int) rep -> 'a vo -> 
            ('a,unit) rep
end
\end{code}

The type of our containers includes the lifted domain |Dom| as one of
the components. This is quite convenient since operations on
containers are usually accompanied by the operations on the retrieved
and stored values. The particular instances of the containers are
parametric over a |DOMAINL|, i.e., functors from a |DOMAINL| module to
the actual implementation of a container.  For example, the following
functor defines a matrix container as a single array, with elements
stored in the row-major order -- the container used in
Fig.~\ref{fig:GE-algorithm}.
\begin{code}[commandchars=\\\{\}]
module GenericVectorContainer(Dom:DOMAINL) =
  struct
  module Dom = Dom
  type contr = Dom.v container2dfromvector
  type 'a vc = ('a,contr) code
  type 'a vo = ('a,Dom.v) code
  let getL x i j = .< ((.~x).arr).(.~i* (.~x).m + .~j) >.
  let dim2 x = .< (.~x).n >.      (* number of rows *)
  let dim1 x = .< (.~x).m >.      (* number of cols *)
  \dots
\end{code}
The accompanying code \cite{metamonadsURL} includes the container with
elements stored in a 1D array in a column-wise (Fortran-like)
mode, and the container for a matrix represented as an array of rows.

We could have defined the type |CONTAINER2D| to be a functor with
|DOMAINL| as an argument. The type |CONTAINER2D| is used in the
signatures of other functors such as |GenLA| of
\S\ref{sec:main-generation} below. If the type |CONTAINER2D| were a
functor, |GenLA| would have been a higher-order functor, and we would
have to pass to |GenLA| two arguments: the container functor and a
|DOMAINL| to apply the container functor to. In the current design,
the user first builds a particular container instance by applying the
functor such as |GenericVectorContainer| to the desired domain. The
user then passes this container instance to |GenLA| as a single
argument. The current design simplifies module signatures at
the expense of making the instantiation of the GE algorithm
`multi-stage'. The stepwise instantiation seems more intuitive
however; it is certainly faster since currently OCaml is quite slow
when instantiating functors with complex signatures.

As mentioned above, the generic GE algorithm |GenLA| is paremeterized
over |CONTAINER2D|.  Given a particular container instance, the
functor |GenLA| yields a module containing various algorithmic aspects
for the user to choose, as well as the main GE driver. The given
container instance is available to all these aspects under the name of
|C| (so that the type of the container can be referred to as |C.contr|
and the dimensions can be obtained using |C.dim1| and |C.dim2|). The
numeric domain, which is part of the container, can be referred to as
|D.Dom| and the type of the domain elements is |C.Dom.v|. We shall see
many such references as we describe particular algorithmic aspects
below.

@@

The determinant aspect is one of several tracking and strategic
aspects. The main GE procedure invokes various functions of these
aspects at some interesting points, e.g., the pivot is required, two
rows have to be permuted, the final answer to the user has to be
built. An aspect may do something at each point, for
example, find pivot according to a pivoting strategy, update the
current value of the determinant, etc.

Our sample Gaussian Elimination algorithm, Fig.~\ref{fig:GE-algorithm},
demonstrates that tracking determinant is complex: first we define the
variables |det_sign| and |det_magn| used for tracking (lines 22 and
23), then we have to change the sign when swapping two rows or two
columns or when the matrix found to be singular (lines 32, 37, 59).
The value of the determinant should be updated per each pivoting (line
56). Finally, we convert the tracking state to the resulting
determinant value (lines 64-66). We observe these lines are not
contiguous: the determinant aspect is consulted at several separate
places in the GE algorithm, and the aspect is supposed to keep
state. As other aspects, the determinant aspect is a module of the
following signature:
\begin{code}
module type DETERMINANT = sig
  type tdet = C.Dom.v ref
  type 'a lstate
  type 'pc pc_constraint = unit
    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
  type ('pc,'v) lm = ('pc,'v) cmonad
    constraint _  = 'pc pc_constraint
  type ('pc,'v) om = ('pc,'v) omonad
    constraint _  = 'pc pc_constraint
  type 'pc nm = ('p,unit) monad
    constraint _ = ('pc,'p,_) cmonad_constraint
    constraint _ = 'pc pc_constraint
  val decl : unit -> 'b nm (* no code is generated *)
  val upd_sign  : unit -> ('b,unit) om
  val zero_sign : unit -> ('b,unit) lm
  val acc_magn  : ('a,C.Dom.v) rep -> (<classif : 'a; ..>,unit) lm
  val get_magn  : unit -> ('b,tdet) lm
  val set_magn  : ('a,C.Dom.v) rep -> (<classif : 'a; ..>,unit) lm
  val fin       : unit -> ('b,C.Dom.v) lm
end
\end{code}
The type |'a lstate| denotes the state being kept as part of the
overall monadic state (tagged as |`TDet|). The signature specifies the
operations of the aspect, such as |decl| for initializing the tracking
state, |zero_sign| to note matrix singularity, |set_magn| to set the
magnitude and |fin| to convert the tracking state to the resulting
value, of  the type |C.Dom.v|, the type of the container
elements. Other aspects follow a similar outline.
\oleg{The following is perhaps too particular to mention in the
  paper. AT least we can keep it in comments:
  We stress the difference in return types of \texttt{decl},
  \texttt{upd\_sign} and \texttt{zero\_sign} functions. The former says
  that \texttt{decl} is an action which is executed only for its
  side-effect. It yields no interpretation object (e.g., produces no
  code value). The function \texttt{zero\_sign} always produces a code
  value -- an expression such as assignment that has the type unit. 
  The function
  \texttt{upd\_sign} is the generator that may produce code, or may
  not. The option type lets us avoid generating code such as 
  \texttt{a := b; ()} with useless dummies.}

@@

For completeness, we briefly describe and show the signatures of the
other aspects. |RANK| tracks rank; this aspect is a simpler
version of determinant tracking. |PIVOTKIND| defines the
representation of the permutation matrix accumulating pivoting
permutations; |TRACKPIVOT| specifies if these permutations are
tracked at all. |LOWER| tracks the |L| factor when the GE algorithm is
used for the in-place LU-decomposition. |PIVOT| is a higher-order
functor abstracting over the pivoting algorithm (full pivoting, row
pivoting, no pivoting, etc). If pivot is found, the aspect should swap
rows and columns appropriately. The swapping may need to be tracked in
the permutation matrix. If we track the determinant,
the swapping must update its sign. The
swapping also affects the |L| factor being accumulated in-place. The
|UPDATE| aspect abstracts over the algorithm for updating matrix
elements during row-reductions: full division update or fraction-free
update. Finally, |INPUT| is an interface aspect letting us handle both
ordinary and augmented matrices.

\begin{code2}
module type RANK = sig
  type 'a tag_lstate
  val decl   : unit -> ('b, int ref) lm
  val succ   : unit -> ('b, unit) lm
  val fin    : unit -> ('b, int) lm
end

@@

The type |wmatrix| is for convenience, a bundling of several pieces of
data needed by both |OUTPUT| and |PIVOT| aspects:
\begin{code}
type 'a wmatrix = {matrix: 'a C.vc; numrow: ('a,int) rep; 
                   numcol: ('a,int) rep}
\end{code}

The module of signature |INTERNAL_FEATURES| bundles information
tracking aspects. The latter are not directly selectable by the
user. Rather, they are \emph{functions} of other user choices. The
implementations of the tracking aspects such as |Rank|, |NoRank|,
|PackedLower| are quite similar (and simpler) than the implementation
of the |AbstractDet| and |NoDet| aspects in \S\ref{sec:det}. The
choice of a particular tracking aspect may depend on all other choices
(e.g., it is not possible to extract the |L| factor if the domain is
not a field). Currently we use preflight checks to ensure consistency
of tracking aspects. Previously~\cite{CaretteKiselyov05} we tried to
implement all our preflight tests at the (module) type level, using
sharing constraints and module computations. That lead to obscure
code, long impenetrable type error messages and very slow compilation.
Furthermore, type-level computations in OCaml are not
powerful enough for the tasks such as verifying that an
integer is prime, see \S\ref{sec:domains}.
